Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
New Abstrac

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction
  for reasoning about program equivalences and deriving optimisations which has
  been successfully applied to functional programming. However, formalising
  this technique in a type theory, (e.g. Coq) often requires compromises, such
  as imposing performance penalties, requiring the assumption of additional
  axioms, or introducing unsafe casts into extracted code (for example,
  \ocaml{Obj.magic} in OCaml).

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion while ensuring termination and providing reasoning about
program equivalences as well as deriving optimisations which has been
successfully applied to functional programming. Formalising recursion schemes in
a type theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (for example, \ocaml{Obj.magic} in OCaml).

This paper presents the first Coq formalisation of a recursion scheme, called
the \emph{hylomorphism}, along with its algebraic laws allowing for the
mechanisation of all recognised (terminating) recursive algorithms. The key
contribution of this paper is that this formalisation is fully axiom-free
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive paradigms such as divide-and conquer, dynamic programming,
and mutual recursion and demonstrate that the extracted OCaml code for
the programs formalised in our framework is efficient, resembles code that a
human programmer would write, and contains no occurrences of \ocaml{Obj.magic}.
We also present a machine-checked proof of the well-known short-cut fusion
optimisation.
\end{abstract}
